MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-06-07 MAESTROeX_I/results.tex
sec 4 readthrough

In the left panel of Figure \ref{fig:scaling} we compare the wallclock time per time step as a function of total core count (in this case, the total number of OpenMP threads) for the original FBoxLib-based MAESTRO implementation to the AMReX MAESTROeX implementation.
These tests were performed using the original temporal integration strategy in \cite{MAESTRO_V}, noting that the new temporal integration with and without the irregular base state gives essentially the same results.
We also include a plot of MAESTROeX without base state evolution.
Comparing the original and new implementations, we see similar scaling results except for the largest simulation, where MAESTROeX performs better.
We see that the increase in wallclock time from the smallest to largest simulation is roughly 42\%.
We also note that without base state evolution, the code runs 14\% faster for small problems, and scale mush better
However, for the case without base state evolution, the overall runtime is faster and the increase in wallclock time from the smallest to largest simulation is roughly 13\%.
This is quite remarkable since there are 3 linear solver per time step (2 cell-centered Poisson solves used in the MAC projection, and a nodal Poisson solve used to compute the updated cell-centered velocities).
Contrary to our prior assumptions, the linear solves are not the primary scaling bottleneck in this code.
In the right panel of Figure \ref{fig:scaling}, we isolate the wallclock time required for these linear solves and see that (i) the linear solves only use 20-23\% of the total computational time, and (ii) the increase in the solver wallclock time from the smallest to largest simulation is only 28\%.
Further profiling reveals that the primary scaling bottleneck is the average operator.
The averaging operator requires binning the sum of Cartesian data onto one-dimensional arrays holding every possible mapping radius.
This amounts to at least 24,384 double precision values (for the $256^3$ simulation) up to 883,584 values (for the $1536^3$ simulation).
The averaging operator requires a global sum reduction over all processors, and the communication of this data is the primary scaling bottleneck.
For the simulation with base state evolution, this averaging operator is only called once per time step (as opposed to 14 times per time step when base state evolution is included).
The difference in total wallclock times with and without base state evolution is almost entirely due to the averaging.
Note that as expected, advection, reactions, and calls to the equation of state scale almost perfectly, since there is only a single parallel communication call to fill ghost cells.

@@

To explore the accuracy and efficiency of the new temporal algorithm, we now analyze in detail three-dimensional, full-star calculations of convection preceding ignition in a sub-Chandrasekhar mass white dwarf. Again, we refer the reader to \cite{MAESTRO_AMR} and \cite{MAESTRO_convection} for setup details. We implement both uniformly- and irregularly-spaced base state with the new temporal algorithm, while only uniform base state spacing is used in the original algorithm. As in Section 3 of \cite{MAESTRO_AMR}, we choose the peak temperature and peak Mach number as the two diagnostics to compare the simulations. Figure \ref{fig:wdconvect_256_maxvar} shows the evolution of both peak temperature and peak Mach number until time of ignition on a single-level grid with resolution of $256^3$. The simulation using the new temporal scheme with uniformly-spaced base state gives the same qualitative results as the original scheme, and predicts a similar time of ignition ($t=7810$ s compared to $t=7850$ s for original algorithm). The simulation using the new temporal scheme with irregularly-spaced base state displays a slightly different peak temperature behavior during the initial transition period $t<150$ s, which results in the difference between the curves post transition. We strongly suspect that this is a result of using different initial model inputs for the star in which the resolution near the center of the star is much coarser with the irregular spacing than the uniform spacing. Fortunately, the simulation with irregular base state spacing still follows the same trend as with uniform spacing, and the star is shown to ignite at an earlier time $t=6840$ s.

Figure \ref{fig:wdconvect_Tmax} shows the peak temperature evolution over the first 1000 s on two grids of differing resolutions, $256^3$ and $512^3$. As previously suspected, the simulation using irregularly-spaced base state agrees much closer to the results from using uniform spacing as the resolution is increased. This is most likely due to the increased resolution of the initial star model for the irregular-spaced base state that is now able to properly capture the same properties near the center of the star as its uniformly-spaced counterpart. This is especially important when computing the base state pressure from base state density since we enforce a constant base state pressure outside of a prescribed cutoff density radius that could potentially be quite different with the two types of base state spacing.

@@

In terms of efficiency, all three simulations on $256^3$ single-level grid were run on Cori haswell with 64 processors and 8 threads per core. The simulation using the new temporal algorithm took 6.75 s per time step with uniformly-spaced base state, which is 13\% faster than the 7.77 s per time step when using the original scheme, and 9.72 s with irregularly-spaced base state. The 25\% increase in run time of the irregularly-spaced algorithm is mainly caused by the much larger base state array and could be significantly reduced if we consider truncating the base state beyond the cutoff density radius.

@@

We now test the performance of MAESTROeX for adaptive, three-dimensional simulations to track a localized feature over time. Figure \ref{fig:amr_grids} illustrates the initial grid structures with two levels of refinement for both planar and spherical geometries where the grid is refined according to the temperature and density profiles, respectively. For each of the problems we tested, the single-level simulation was run using the original temporal scheme and the adaptive simulations using the original and new temporal algorithms. We want to show that the adaptive simulation can give the similar results as the single-level simulation and in a more computationally efficient manner.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=2.5in]{./figs/reacting_bubble_result} \hspace{2.5em}
\includegraphics[width=2.5in]{./figs/reacting_bubble_amr_result}
\caption{\label{fig:bubble_results} Time-lapse cross-section of a hot bubble in a white dwarf environment at 
         $t = 0$, 1.25, and 2.5 s for (left) single-level simulation, and 
         (right) adaptive simulation at the same effective resolution.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In the left panel of Figure \ref{fig:scaling} we compare the wallclock time per time step as a function of total core count (in this case, the total number of OpenMP threads) for the original FBoxLib-based MAESTRO implementation to the AMReX MAESTROeX implementation.
These tests were performed using the original temporal integration strategy in \cite{MAESTRO_V}, noting that the new temporal integration with and without the irregular base state gives essentially the same results.
We also include a plot of MAESTROeX without base state evolution.
Comparing the original and new implementations, we see similar scaling results except for the largest simulation, where MAESTROeX performs better.
We see that the increase in wallclock time from the smallest to largest simulation is roughly 42\%.
We also note that without base state evolution, the code runs 14\% faster for small problems, and scales much better with wallclock time from the smallest to largest simulation increasing by only 13\%.
This is quite remarkable since there are 3 linear solver per time step (2 cell-centered Poisson solves used in the MAC projection, and a nodal Poisson solve used to compute the updated cell-centered velocities).
Contrary to our prior assumptions, the linear solves are not the primary scaling bottleneck in this code.
In the right panel of Figure \ref{fig:scaling}, we isolate the wallclock time required for these linear solves and see that (i) the linear solves only use 20-23\% of the total computational time, and (ii) the increase in the solver wallclock time from the smallest to largest simulation is only 28\%.
Further profiling reveals that the primary scaling bottleneck is the average operator.
The averaging operator requires collecting the sum of Cartesian data onto one-dimensional arrays holding every possible mapping radius.
This amounts to at least 24,384 double precision values (for the $256^3$ simulation) up to 883,584 values (for the $1536^3$ simulation).
The averaging operator requires a global sum reduction over all processors, and the communication of this data is the primary scaling bottleneck.
For the simulation with base state evolution, this averaging operator is only called once per time step (as opposed to 14 times per time step when base state evolution is included).
The difference in total wallclock times with and without base state evolution is almost entirely due to the averaging.
Note that as expected, advection, reactions, and calls to the equation of state scale almost perfectly, since there is only a single parallel communication call to fill ghost cells.

@@

To explore the accuracy of the new temporal algorithm, we now analyze in detail three-dimensional, full-star calculations of convection preceding ignition in a sub-Chandrasekhar mass white dwarf. Again, we refer the reader to \cite{MAESTRO_AMR} and \cite{MAESTRO_convection} for setup details. We implement both uniformly- and irregularly-spaced base state with the new temporal algorithm, while only uniform base state spacing is used in the original algorithm. As in Section 3 of \cite{MAESTRO_AMR}, we choose the peak temperature and peak Mach number as the two diagnostics to compare the simulations. Figure \ref{fig:wdconvect_256_maxvar} shows the evolution of both peak temperature and peak Mach number until time of ignition on a single-level grid with resolution of $256^3$. The simulation using the new temporal scheme with uniformly-spaced base state gives the same qualitative results as the original scheme, and predicts a similar time of ignition ($t=7810$ s compared to $t=7850$ s for original algorithm). The simulation using the new temporal scheme with irregularly-spaced base state displays a slightly different peak temperature behavior during the initial transition period $t<150$ s, which results in the difference between the curves post transition. We strongly suspect that this is a result of using different initial model inputs for the star in which the resolution near the center of the star is much coarser with the irregular spacing than the uniform spacing. Fortunately, the simulation with irregular base state spacing still follows the same trend as with uniform spacing, and the star is shown to ignite at an earlier time $t=6840$ s.

Figure \ref{fig:wdconvect_Tmax} shows the peak temperature evolution over the first 1000 s on two grids of differing resolutions, $256^3$ and $512^3$. Limited allocations prevented us from running this simulation further.
As previously suspected, the simulation using irregularly-spaced base state agrees much closer to the results from using uniform spacing as the resolution is increased. This is most likely due to the increased resolution of the initial star model for the irregular-spaced base state that is now able to properly capture the same properties near the center of the star as its uniformly-spaced counterpart. This is especially important when computing the base state pressure from base state density since we enforce a constant base state pressure outside of a prescribed cutoff density radius that could potentially be quite different with the two types of base state spacing at lower resolutions.

@@

In terms of efficiency, all three simulations on $256^3$ single-level grid were run on Cori haswell with 64 processors and 8 threads per core. The simulation using the new temporal algorithm took 6.75 s per time step with uniformly-spaced base state, which is 13\% faster than the 7.77 s per time step when using the original scheme, and 9.72 s with irregularly-spaced base state. The 25\% increase in run time of the irregularly-spaced algorithm is mainly caused by the much larger base state array and could be significantly reduced if we consider truncating the base state beyond the cutoff density radius. 
\MarginPar{why is the new temporal with constant dr faster?}

@@

We now test the performance of MAESTROeX for adaptive, three-dimensional simulations to track localized regions of interest over time. Figure \ref{fig:amr_grids} illustrates the initial grid structures with two levels of refinement for both planar and spherical geometries where the grid is refined according to the temperature and density profiles, respectively. For each of the problems we tested, the single-level simulation was run using the original temporal scheme and the adaptive simulations using the original and new temporal algorithms. We want to show that the adaptive simulation can give the similar results as the single-level simulation and in a more computationally efficient manner.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=2.5in]{./figs/reacting_bubble_result} \hspace{2.5em}
\includegraphics[width=2.5in]{./figs/reacting_bubble_amr_result}
\caption{\label{fig:bubble_results} Time-lapse cross-section of a hot bubble in a white dwarf environment at 
         $t = 0$, 1.25, and 2.5 s for (left) single-level simulation, and 
         (right) adaptive simulation at the same effective resolution.  The red, green, and black boxes indicate grids of increasing resolution}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
